FLM for growth as function of SPEI

### 2021-02-22 analysis/paper/EDA_figure_1.Rmd
updated figures and results
updates to results

@@

```{r}
demog_post %>% ggplot(aes(x = habitat, y = log_size)) + geom_boxplot()
```
```{r}
demog_post %>% 
  group_by(habitat) %>% 
  summarize(size = mean(size, na.rm = TRUE),
            shts = mean(shts, na.rm = TRUE),
            ht = mean(ht, na.rm = TRUE))
```

@@

```{r}
size <-
  demog_plotdf %>% 
  ggplot(aes(x = yearmonth, y = size, color = habitat)) +
  stat_summary(geom = "line", fun = "mean", fun.args = list(na.rm = TRUE)) +
  stat_summary(geom = "pointrange", fatten = 1) +
  scale_x_yearmonth(limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
  scale_y_continuous("Size") +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "none"
  )
size
# size/survival/spei
```
Average plant size shrank in 2003 drought, but not 2007 drought

>>>>

When is survival lowest?

```{r}
demog_post %>% 
  group_by(year, habitat) %>% 
  summarize(surv = sum(surv, na.rm = TRUE) / n()) %>% 
  pivot_wider(names_from = habitat, values_from = surv)
```

@@

```{r}
demog_post %>% ggplot(aes(x = habitat, y = log_size)) + geom_boxplot()
```
```{r}
demog_post %>% 
  group_by(habitat) %>% 
  summarize(across(
    c(size, shts, ht),
    list(mean = mean, sd = sd),
    na.rm = TRUE,
    .names = "{.col}_{.fn}"
  ))

```

@@

```{r}
size <-
  demog_plotdf %>% 
  ggplot(aes(x = yearmonth, y = log(size), color = habitat)) +
    stat_summary(
      geom = "line",
      fun = "mean",
      fun.args = list(na.rm = TRUE),
      position = position_dodge(width = 100),
      aes(group = habitat)
    )+
    stat_summary(
      geom = "pointrange",
      fatten = 1,
      fun.data = "mean_sdl",
      fun.args = list(mult = 1),
      position = position_dodge(width = 100)
    ) +
  scale_x_yearmonth(limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
  scale_y_continuous("log(size)") +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "none"
  ) +
  NULL
# size/survival/spei
```
Average plant size shrank in 2003 drought, but not 2007 drought

### Ridge plot
```{r}
library(ggridges)
ggplot(demog_plotdf) +
  geom_density_ridges(aes(x = log(size), y = as.factor(year), fill = habitat, color = habitat), alpha = 0.5)
ggplot(demog_plotdf) +
  geom_violin(aes(y = log(size), x = as.factor(year), fill = habitat), alpha = 0.5, draw_quantiles = c(0.5))

```

### 2021-02-22 analysis/paper/paper.Rmd
updates to results

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
library(knitr)
library(here)
```

@@

The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
In addition to large areas of continuous forest, the BDFFP has forest fragment reserves that range in size from 1-100 ha (Figure \@ref(fig:map)). These fragments were isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and - in most cases - burning the downed trees once they dried. In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].

@@

Precipitation and potential evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
SPEI can be calculated on different scales and we chose to calculate 3-month SPEI because *H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (Appendix 1).

@@

```{r}
cap <- "Timeseries of demographic parameters and drought ocurrence. Mean plant size ± standard error (a) shows that plants were smaller in fragments (red) than continuous forest (blue). (b) shows the survival curve for the 1998 cohort of plants in both fragments and continuous forest. c) shows the proportion of mature plants (with 4 or more shoots) flowering each census. In d), monthly 3-month SPEI is plotted. Light grey lines represent values from different grid cells encompassing BDFFP and the dark line is the site mean.  Yellow, orange, dark orange, and red stripes show mild, moderate, severe, and extreme drought, respectively."
```

@@

The meteorological drought occurrences indicated by SPEI mostly agree with drought events in the Amazon reported in the literature.
For example, the 1997 El Niño drought was one of the most severe droughts in the Amazon on record [@williamsonAmazonianTreeMortality2000] and correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).\
The 2005 dry season (June--October) was also reported as exceptionally dry, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
<!--# I know I'm missing some literature here.  Add more refs for more droughts. Any papers mentioning 2003 wet season?-->

@@

@@

Plants were on average larger in continuous forest compared to 1-ha fragments.
In continuous forest, plants had an average of 2.91 shoots and were on average 40.6 cm tall with an average size (shoots $\times$ height) of 150.
In 1-ha fragments, plants had 2.53 shoots on average and a mean height of 36.2 cm and a mean size of 115.
The size difference between plants in continuous forest and 1-ha fragments was more dramatic early on in the timeseries with mean sizes becoming more similar over time (Figure \@ref(fig:eda)a).

@@

<!--# There's a lot more going on in the 1-ha fragments.  Hurts my brain to look at it.  Will write another day. -->

@@

Plant size had a significant effect on flowering probability in both habitats (p \< 0.001).
Flowering probability increased rapidly with log-transformed plant size greater than 5 in both habitats (Figure \@ref(fig:flwr-size)).
Flowering probability was greater in 1-ha fragments for large plants compared to continuous forest.
However, because mean plant size is larger in continuous forest, overall flowering probability was greater.
<!--# idea: plot vertical lines representing mean or median size in each habitat -->

@@

>>>>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "../figures/",
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>",
	dpi = 300
)
library(knitr)
library(here)
```

@@

The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
In addition to large areas of continuous forest, the BDFFP has forest fragment reserves that range in size from 1-100 ha (Figure \@ref(fig:map)). These fragments were isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried. In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].

@@

Precipitation and potential evapotranspiration data were obtained from a published gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016]. We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
SPEI can be calculated to represent different temporal scales of drought.
We used 3-month SPEI because *H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (**Appendix \_\_**).

@@

```{r}
cap <- "Timeseries of demographic parameters and drought ocurrence. Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (red) than continuous forest (blue). (b) shows the survival curve for the 1998 cohort of plants in both fragments and continuous forest. c) shows the proportion of mature plants (with 4 or more shoots) flowering each census. In d), monthly 3-month SPEI is plotted. Light grey lines represent values from different grid cells encompassing BDFFP and the dark line is the site mean.  Yellow, orange, dark orange, and red stripes show mild, moderate, severe, and extreme drought, respectively."
```

@@

The meteorological drought occurrences indicated by SPEI mostly agree with drought events in the Amazon reported in the literature.
For example, the 1997 El Niño drought was one of the most severe droughts in the Amazon on record [@williamsonAmazonianTreeMortality2000] and correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).\
The 2005 dry season (June--October) was also reported as exceptionally dry, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
<!--# I know I'm missing some literature here.  Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->

@@

```{r}
cap <- "The difference between predicted effects of SPEI history on survival in continuous forest and 1-ha fragments."
```

```{r surv-diff, fig.cap=cap}
include_graphics(here("analysis", "figures", "surv_diff_spei.png"))
```

Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-diff)).
However, under extreme conditions, survival probability is higher in continuous forest by up to 0.02.

@@

Plants were on average larger in continuous forest compared to 1-ha fragments.
<!--# This is across the whole timeseries, which I'm not sure is actually that important. What summary statistics would be useful? -->In continuous forest, plants had an average of 2.9 ± 1.8 shoots (mean ± standard deviation) and were on average 40.6 ± 26.5 cm tall with an average size (shoots $\times$ height) of 150 ± 175.
In 1-ha fragments, plants had 2.5 ± 1.5 shoots on average and a mean height of 36.2 ± 24.1 cm and a mean size of 112 ± 141.
The size difference between plants in continuous forest and 1-ha fragments was more dramatic early on in the timeseries with mean sizes becoming more similar over time (Figure \@ref(fig:eda)a).

@@

```{r}
cap <- "The difference between predicted effects of SPEI history on plant size in continuous forest and 1-ha fragments."
```

```{r growth-diff, fig.cap=cap}
include_graphics(here("analysis", "figures", "growth_diff_spei.png"))
```

The difference between the effects of SPEI history in continuous forest and 1-ha fragments shows that growth is equal or slightly higher in continuous forest for average conditions (SPEI = 0) over all lag periods (Figure \@ref(fig:growth-diff).
However, growth in continuous forests is higher by up to log(size) = 0.48 <!--# double-check this number --> when the current wet season is unusually wet (SPEI\>1).

@@

Plant size had a significant effect on flowering probability in both habitats (p \< 0.001).
Flowering probability increased rapidly with log-transformed plant size greater than 5 in both habitats (Figure \@ref(fig:flwr-size)).
Flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest.
However, because mean plant size is larger in continuous forest, overall flowering probability was greater.
<!--# idea: plot vertical lines representing mean or median size in each habitat -->

@@

```{r}
cap <- "The difference between predicted effects of SPEI history on flowering probability in continuous forest and 1-ha fragments."
```

```{r flwr-diff, fig.cap=cap}
include_graphics(here("analysis", "figures", "flwr_diff_spei.png"))
```

Plotting the difference between the crossbasis smooths for continuous forest and 1-ha fragments revealed that flowering probability is higher in continuous forest compared to 1-ha fragments and is increased by recent drought and drought during the dry season 15--20 months prior to the census (Figure \@ref(fig:flwr-diff)).
The least difference between habitat types occurs when there is strong drought at about 8--12 months lag or very wet conditions at 15--20 months lag.

### 2021-02-22 analysis/paper/vital rate models.Rmd
updates to results

```{r}
f_diff_rotated <- 
  bind_cols(
  f_cf_eval %>% rename_with(.fn = ~glue("cf_{.}")),
  f_1ha_eval %>% rename_with(.fn = ~glue("frag_{.}"))
  ) %>%
  mutate(diff = cf_est - frag_est) %>% 
  ggplot(aes(x = cf_L, y = cf_spei_history, fill = diff, z = diff)) +
  geom_raster() +
  geom_contour(color = "black", binwidth = 0.001) +
  geom_hline(aes(yintercept = 0), color = "grey", linetype = 2) +
  scale_fill_viridis_c("∆Flowering (CF – 1ha)", option = "viridis") +
  scale_x_continuous(
    "lag (months before census)",
    breaks = breaks,
    expand = c(0, 0)
  )+
  scale_y_continuous(TeX("SPEI_3_"), expand = expansion(mult = c(0.025, 0))) +
  labs(title = "CF – 1-ha") +
  theme_bw() +
  annotation_custom(grob = ggplotGrob(season_bar), ymin = -Inf, ymax = min(s_cf_eval$spei_history), xmin = -Inf, xmax = Inf) +
  NULL
ggsave(here("analysis", "figures", "flwr_diff_spei.png"),plot = s_diff_rotated, width = 7, height = 2.5)
```

>>>>

```{r}
f_diff_rotated <- 
  bind_cols(
  f_cf_eval %>% rename_with(.fn = ~glue("cf_{.}")),
  f_1ha_eval %>% rename_with(.fn = ~glue("frag_{.}"))
  ) %>%
  mutate(diff = cf_est - frag_est) %>% 
  ggplot(aes(x = cf_L, y = cf_spei_history, fill = diff, z = diff)) +
  geom_raster() +
  geom_contour(color = "black", binwidth = 0.001) +
  geom_hline(aes(yintercept = 0), color = "grey", linetype = 2) +
  scale_fill_viridis_c("∆Flowering (CF – 1ha)", option = "viridis") +
  scale_x_continuous(
    "lag (months before census)",
    breaks = breaks,
    expand = c(0, 0)
  )+
  scale_y_continuous(TeX("SPEI_3_"), expand = expansion(mult = c(0.025, 0))) +
  labs(title = "CF – 1-ha") +
  theme_bw() +
  annotation_custom(grob = ggplotGrob(season_bar), ymin = -Inf, ymax = min(s_cf_eval$spei_history), xmin = -Inf, xmax = Inf) +
  NULL
ggsave(here("analysis", "figures", "flwr_diff_spei.png"),plot = f_diff_rotated, width = 7, height = 2.5)
```
